---
title: "Chromosome compaction - simple three-state model"
output: 
  html_notebook:
    toc: false
    toc_float: false
    code_folding: hide
    css: nice_notebook.css
---

```{r}
library(mylib)
library(ChromCom)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(parallel)

binDir <- "../RData"
```

This is a continuation of the previous notebook, [mod3_notes_old](), with a new model. In the old model, the start time $t_1$ was a fixed number, creating a cusp in the curve, while in the new model is a random variable with exponential distribution.

# 17/08/2017

The new model. An example for $tau = 20$ min, $\Delta t = 0$, $k_1 = 0.04\ {\rm min}^{-1}$ and $k_2 = 0.04\ {\rm min}^{-1}$.

```{r, fig.width=5, fig.height=3}
pars <- c3pars(
  tau = 20,
  dt2 = 0,
  k1 = 0.04,
  k2 = 0.04
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=1000)
plotTimelines(chr)
```

## Experimental data

```{r, warning=FALSE, fig.width=5}
echr.scr <- experimentalData(dataFile$scramble)
echr.d2 <- experimentalData(dataFile$NCAPD2)
echr.d3 <- experimentalData(dataFile$NCAPD3)
echr.smc <- experimentalData(dataFile$SMC2)

plotTimelines(echr.scr, smooth=TRUE, k=15)
```



I start with fitting scramble

## Scramble

Fitting with $k_1$, $k_2$ and $tau$ free.

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2")
chr.scr <- cacheData("fitv_scramble_n3000_tau_k1_k2", fitChr, echr.scr, pars, free, nsim=3000, ntry=32, ncores=6, binDir=binDir)
plotTimelines(chr.scr, expdata = echr.scr, withpars=TRUE)
```

It looks like we need to free $\Delta t_2$.

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2", "dt2")
chr.scr.d2 <- cacheData("fitv_scramble_n3000_tau_k1_k2_dt2", fitChr, echr.scr, pars, free, nsim=3000, ntry=32, ncores=6, binDir=binDir)
plotTimelines(chr.scr.d2, expdata = echr.scr, withpars=TRUE)
```
