---
title: "Chromosome compation - simple 3-state model"
author: "Marek Gierlinski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: true
---


```{r setup, include=FALSE}
library(ChromCom)
library(knitr)
library(mylib)
library(gridExtra)
binDir <- "../RData"
opts_chunk$set(fig.path='../figure/mpd3-', 
               cache.path='../cache/mod3-', 
               fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.pos='H',
               fig.width=4,
               fig.height=3
              )
```

# Data

```{r}
echr <- experimentalData(dataFile$scramble)
smth <- 5
```

In our data we distinguish four states, marked by colour: blue, brown, pink, red. For the purpose of the model we merge blue and brown together into one state. Here is the result of the experiment from `r nrow(echr$cells)` cells.

```{r}
plotCells(echr)
```

The next figure shows the proportions of each colour as a function of time. The curves are smoothed with a running mean over the window of `r smth` time points.

```{r}
plotTimelines(echr, smooth=TRUE, k=smth)
```


# Model

The model consists of three states and a set of rules.

## States

* blue/brown (B)
* pink (P)
* red (R)
 
## Rules

- Simulation is carried out at a discrete time step of 1 min
- The cell is in state B before time $t_1$.
- B$\rightarrow$P occurs after $t_1$ with rate $k_1$
- P$\rightarrow$R occurs after $t_1 + \Delta t_2$ with rate $k_2$
- P$\rightarrow$B occurs after $t_1 + \Delta t_3$ with rate $k_3$

## Details

There are two ways the model can be calculated. The transition model is simpler, but does not allow for P$\rightarrow$B transitions. The simulation model is more flexible and, at this moment, applied as default.

### "Transition" model

For each cell we generate two transitions times (B$\rightarrow$P and P$\rightarrow$R) according to an exponential distribution, 

\[
\begin{eqnarray}
Pr(B \rightarrow P; t) & = & 1 - e^{-k_1 (t - t_1)} \\
Pr(P \rightarrow R; t) & = & 1 - e^{-k_2 \left[t - (t_1 + t_{\rm BP} + \Delta t_2\right]}
\end{eqnarray}
\]

where $t_{\rm BP}$ is the B$\rightarrow$P transition time, generated with the first distribution. The exponential distribution describes a process with constant probability of the event happening. From transition times we build the timeline of the cell, its state as a function of time. 

### "Simulation" model

This is a Markov chain approach. The next state is generated from the current state based on rules outlined above. The rates, $k$, are converted into probabilities over a given time step $\Delta t$ as $Pr = 1 - e^{1 - k\Delta t}$.

With either approach the cell timeline is repeated for 3000 times and then the colour proportions are found at each time point.

## Example

Here is an example of the model.

```{r, fig.width=4, fig.height=3}
pars <- c3pars(
  t1 = -30,
  dt2 = 0,
  k1 = 0.05,
  k2 = 0.03
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=3000)
plotTimelines(chr, withpars=TRUE)
```

## Parameter tuner

There is a [shiny app](https://shiny.compbio.dundee.ac.uk/marek_chromcom/param_tuner/) that allows tuning parameters in search for the best solution.

# Fitting model to the data

I attempted fitting model to our data. It is not an easy task, as the model is stochastic. After some testing I found that modified BFGS (a quasi-Newton method, also known as a variable metric algorithm, by Broyden, Fletcher, Goldfarb and Shanno, 1970) which allows box constraints (Byrd et. al. 1995), gives reasonable results. Again, due to stochasticity of the model, this algorithm often finds false local minima. I run it several times to find the best minimum. It is a crude and time-consuming method, but gives results better than manual tuning.

Fitting is constrained to time points between -90 and 30 min. The minimized quantity is

${\rm rms} = \sqrt{\sum_{c \in \{B,P,R\}} \sum_i (O_{c,i} - E_{c,i})^2}$

that is, the square root of the sum of squared residuals over all time points and all colours.

We apply the following strategy:

- fit scramble with $t_1$, $k_1$ and $k_2$ free
- for NCAPD2 we expect only $k_2$ to change, so fix $k_1$ from scramble
- for NCAPD3 we expect only $k_1$ to change, so fix $k_2$ from scramble; however, John tried this manually and it didn't work very well, so we might free both ks
- for SMC2 $k_1$ fixed from NCAPD3, $k_2$ fixed from NCAPD2 and see if this matches; then do a fit
- lets keep all $\Delta t$ fixed

## Scramble

```{r, fig.width=5}
pars <- c3pars()
free <- c("t1", "k1", "k2")
chr.scr <- cacheData("fit_scramble_n3000_t1_k1_k2", fitChr, echr, pars, free, nsim=3000, ntry=16, ncores=8, binDir=binDir)
plotTimelines(chr, expdata = echr, withpars=TRUE)
```

## NCAPD2

```{r, fig.width=5}
pars <- chr.scr$pars
free <- c("t1", "k2")
chr.d2 <- cacheData("fit_NCAPD2_n3000_t1_k2", fitChr, echr, pars, free, nsim=3000, ntry=16, ncores=8, binDir=binDir)
plotTimelines(chr, expdata = echr, withpars=TRUE)
```

