---
title: "Chromosome compaction - simple three-state model"
output: 
  html_notebook:
    toc: false
    toc_float: false
    code_folding: hide
    css: nice_notebook.css
---

```{r}
library(mylib)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(parallel)
library(latex2exp)

source("../R/lib.R")
binDir <- "../RData"
```


```{r, warning=FALSE}
echr <- lapply(dataFile, experimentalData)
```

This is a continuation of the previous notebook, [Notes.2](), with a new model.


# 31/08/2017

Implementing changes as outlined in the discussion. Now we can force pink to peak before zero, by tweaking the $t_0$ parameter.

```{r, fig.width=5, fig.height=3}
pars <- c3pars(
  t0 = -10,
  tau1 = 10,
  tau2 = 10,
  k1 = 0.05,
  k2 = 0.04
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=1000)
plotTimelines(chr)
```


## Scramble

I did try to fit with five parameters (including $t_0$), but `optim` crashed sooner or later with uncomprehensible error messages. We need to fix $t_0$.

```{r}
pars <- c3pars(
  t0 = -10,
  tau1 = 15.3,
  tau2 = 9,
  k1 = 0.047,
  k2 = 0.064
)
free <- c("tau1", "k1", "k2", "tau2")
chr.scr.10 <- cacheData("fitt_scramble_n10000_t.10_tau1_k1_k2_tau2", fitChr, echr$scr, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)

pars$t0 <- -5
chr.scr.5 <- cacheData("fitt_scramble_n10000_t.5_tau1_k1_k2_tau2", fitChr, echr$scr, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)

pars$t0 <- -15
chr.scr.15 <- cacheData("fitt_scramble_n10000_t.15_tau1_k1_k2_tau2", fitChr, echr$scr, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)

pars$t0 <- 0
chr.scr.0 <- cacheData("fitt_scramble_n10000_t.0_tau1_k1_k2_tau2", fitChr, echr$scr, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)
```


```{r, fig.width=5}
plotTimelines(chr.scr.0, expdata = echr$scr, withpars=TRUE)
plotTimelines(chr.scr.5, expdata = echr$scr, withpars=TRUE)
plotTimelines(chr.scr.10, expdata = echr$scr, withpars=TRUE)
plotTimelines(chr.scr.15, expdata = echr$scr, withpars=TRUE)
```
# 04/09/2017

Need to do this on the cluster. Made a little script `fitting.Rs` to fit one set. Lets try is.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.R"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"

for(set in names(dataFile)) {
  name <- paste0("fitt_", set, "_t0_n100000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name)
  qsub(cmd, logDir, name=name)
}
```

# 05/09/2017

Fit results are ready. Lets have a look

```{r eval=FALSE}
nm <- names(dataFile)
nm <- nm[which(nm != "MK1775_ICRF193")]
chr <- lapply(nm, function(set) {
  name <- paste0("fitt_", set, "_t0_n10000_tau1_k1_k2_tau2")
  cacheData(name, fitChr, binDir=binDir, cacheonly=TRUE)
})
names(chr) <- names(dataFile)
```


```{r eval=FALSE, fig.width=6, fig.height=4}
for(name in names(dataFile)) {
  print(plotTimelines(chr[[name]], expdata = echr[[name]], withpars=TRUE, title=name))
}
```


There are serious issues with data file MK1775andICRF193_noncoloured.csv. First, there are strings "ICRF" in it. Second, there seem to be 6 empty columns at the end. It needs redoing.


# 15/09/2017

The `MK1775andICRF193` file is now corrected. Fit it.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.Rs"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"

set <- "MK1775_ICRF193"
name <- paste0("fitt_", set, "_t0_n100000_tau1_k1_k2_tau2")
cmd <- paste("Rscript", script, set, name)
qsub(cmd, logDir, name=name)
```

## Meeting with Tomo and John

- change slider rates - upper limit to 0.5
- change font size so rms is visible
- fit range from -50 to +30
- modify model: add a switch $t_2 = t_1 + R(\tau_2)$ or $t_2 = t_0 + R(\tau_2)$; this can help with red curves that seem to all go up at 0, for all conditions
- later: fit by changing only one param from data to data


The model is changed. See the switch difference.

```{r, fig.width=5, fig.height=3}
pars <- c3pars(
  t0 = 0,
  tau1 = 40,
  tau2 = 40,
  k1 = 0.05,
  k2 = 0.04,
  t2ref = 1
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=1000)
plotTimelines(chr, withpars = TRUE)
```

```{r, fig.width=5, fig.height=3}
pars <- c3pars(
  t0 = 0,
  tau1 = 40,
  tau2 = 10,
  k1 = 0.02,
  k2 = 0.04,
  t2ref = 0
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=1000)
plotTimelines(chr, withpars=TRUE)
```

A quick scramble fit with the new switch.

```{r}
pars <- c3pars(
  t0 = -10,
  tau1 = 120,
  tau2 = 4,
  k1 = 0.02,
  k2 = 0.064,
  t2ref = 0
)
free <- c("tau1", "k1", "k2", "tau2")
chrs.scr <- fitChr(echr$scr, pars, free, nsim=100, ntry=10, ncores=6)
```

```{r, fig.width=5}
plotTimelines(chrs.scr, expdata = echr$scr, withpars=TRUE)
```





Submitting jobs to the cluster and naively expecting for them not to crash over the weekend. Since the t2ref=0 model look dodgy, I only run t2ref=1.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.R"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 1

for(set in names(dataFile)) {
  name <- paste0("fits_", set, "_ref1_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, 0)
  qsub(cmd, logDir, name=name, ncor = 8)
}
```

# 18/09/2017

There were some problems with fit limtis. I think I should fit it again.

Struggling with TeX expression and multi-line title. Reverting to simple text.

# 19/09/2017

Doing interviews today, so very little time. Trying running the new switch model.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.Rs"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 0

for(set in names(dataFile)) {
  name <- paste0("fits_", set, "_ref0_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, 0)
  qsub(cmd, logDir, name=name)
}
```

Looking at data I have an impression that red curves start a bit earlier than 0. Try the same model but $t_0 = -10$.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.Rs"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 0

for(set in names(dataFile)) {
  name <- paste0("fits_", set, "_ref0_t10_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, -10)
  qsub(cmd, logDir, name=name)
}
```

# 20/09/2017

The last batch of fits crashed. Inside mclapply, that is in the optim, but don't know the error.

It seems text-numeric type conflicts. I explicitly convert parameters into numeric now and fits are OK. Running them again, deeper.

# 25/09/2017

After a few days off work due to illnes, I'm putting all of this together in the mod3 document. The original model (ref1) needs to be re-fitted on the smae time interval as the new one. Running fits again.

# 9/10/2017

Extended model for display (not for rms) in the Shiny app.

# 20/11/2017

Adding two new data files. Need to update `experimentalData` function as it did not work with data file with an empty cell (converted it quietly into NAs).


# 8/01/2018

Looking for ways of finding confidence intervals on fit parameters. Not much found. Asked question on StackOverflow.

Fitting new data sets.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.Rs"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 0

host <- "login-gjb-1.compbio.dundee.ac.uk"

for(set in c("TT103", "TT108")) {
  name <- paste0("fits_", set, "_ref0_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, 0)
  qsub(cmd, logDir)
  
  name <- paste0("fits_", set, "_ref0_t10_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, -10)
  qsub(cmd, logDir)
}

switch <- 1
for(set in c("TT103", "TT108")) {
  name <- paste0("fits_", set, "_ref1_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, 0)
  qsub(cmd, logDir)
}

switch <- 0
for(set in names(dataFile)[1:9]) {
  name <- paste0("fits_", set, "_ref0_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 100, 0)
  qsub(cmd, logDir)
}
```


## Meeting Tomo and John

Distribution of colour length: if Poisson process, then it should be exponential. test in different windows. Early window (<-20 min), see how brown behaves.

New data set: early G2. Late G2 is our data up to -20 min.

Look at density distribution of brown.

Distribution of brown  - are they Poisson, ranodm, or is there a pattern, like clumping. Do cumulative dist. on intervals and see if it is exponential.

How to correct for the lenght of cell available (that is brow-to-brown distance will be biased for short cells, longer intervals missing)?

# 9/01/2018

Working on brown distribution.

# 10/01/2018

Now I know how to find confidence intervals on the parameter fits. Thanks to StackOverflow. We need to use a bootstrap.

- Select random samples, with replacement, from original (x, y) data.
- Fit each of the samples and find bootstrap parameters, tau1*, k1*, k2*
- Repeat 1000 times
- For each parameter find 95% confidence intervals

# 11/01/2018

Trying to improve fitting process, to make it a bit faster. Instead of simulation, I try again the transition method again, but cannot get consistent results.

# 12/01/2018

What if I parallelize cell generation?


```{r test_generate_cells}
# the original code
generateCells1 <- function(chr, nsim=1000, mode="simulation") {
  cells <- t(replicate(nsim, timelineCell(chr$pars, chr$timepars, mode)))
  cnt <- cellCount(cells, chr$time, chr$colours)
  chr$cells <- cells
  chr$cnt <- cnt
  chr$nsim <- nsim

  return(chr)
}

# parallel version
generateCells2 <- function(chr, nsim=1000, mode="simulation", ncores=8) {
  tc <- mclapply(1:nsim, function(i) {
    timelineCell(chr$pars, chr$timepars, mode)
  }, mc.cores = ncores)
  cells <- do.call(rbind, tc) # this is faster
  #cells <- plyr::laply(tc, .fun=identity)
  cnt <- cellCount(cells, chr$time, chr$colours)
  chr$cells <- cells
  chr$cnt <- cnt
  chr$nsim <- nsim

  return(chr)
}

system.time(for(i in  1:10) generateCells1(echr$scramble, nsim=10000))
system.time(for(i in  1:10) generateCells2(echr$scramble, nsim=10000))
```

Parallel version reduces time by half.

## Bootstrap trial

[Moved to main document]



Also, have done quite a lot of work on interval distribution.

# 15/01/2018

Spent a lot of time on brown distribution, in particular gap between gap events. Randomizing data was tricky.


# 16/08/2018

More work on brown distribution. Improving statistical tests. Running parameter bootstraps for other data.

# 17/08/2018

Adding new results, tidying up the document. Also, done chi-square tests on S and G2 data.

# 18/01/2018

Improving fit parameter bootstrap figures.

Meeting with Tomo and John: do point-to-point diagrams for all conditions (also S and G2). Do a chi-square test between these.

# 19/01/2018

Added SMC2. Added point-to-point distributions everwhere.

# 23/01/2018

Running more bootstraps. Yesterday, there were cluster problems.

Meeting with Tomo and John.

Why do model lines in fits always start at -50? That is blue = 1, pink = red = 0, but the model should start earlier.

Brown distribution: if brown points are randomly scattered, what is the prob. of seeing 2, 3, 4... events next to each other. Compare with duration distirbution.

# 24/01/2018

Chasing errors in the fitting code. It turns out I used the RMS window (-50, 30) not only for RMS, but also for the model. This is not right. The model always starts in blue, so needs some time to stabilize. If it starts in blue at -50 mins, then there is a very strong boundary condition. Now I changed it. This is how it works now:

- the model is calculated at the same grid as the data
- time points between -50 and 30 are selected for RMS only
- the selection is done either for bootstrap (sampling) and non bootstrap (-50, -49, ..., 30)
- rms is calculated on this selection

Hence, as intended, bootstrap is done only in the active RMS window, but the model is calculated over the full extent of data.

First, re-running the non-bootstrap fits. I deleted all RData fits, so there is no confusion with older, incorrect models.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.R"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 1

for(set in names(dataFile)) {
  name <- paste0("fits_", set, "_ref1_t0_n10000_tau1_k1_k2_tau2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 30, 0, 4)
  qsub(cmd, logDir, name=name, ncor = 8)
}
```

While this is churning on the cluster, I play with brown/pink distribution.

# 25/01/2018

This is very slow. Now I compute the model on the full time grid and it takes much longer!

In the meantime, I can find best fits for three free parameters. I fix $\tau_2 = 8$ min.

```{r eval=FALSE}
script <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/R/fitting.R"
logDir <- "/cluster/gjb_lab/mgierlinski/projects/chromcomR/log/"
switch <- 1

for(set in names(dataFile)) {
  name <- paste0("fits_", set, "_ref1_t0_n10000_tau1_k1_k2")
  cmd <- paste("Rscript", script, set, name, switch, 10000, 30, 0, 3)
  qsub(cmd, logDir, name=name, ncor = 8)
}
```

# 26/01/2018

Running more bootstraps. Hopefully, these will be finished over the weekend.

Single fits for three parameters are ready. They look similar to four-parameter fits.

# 29/01/2018

Continuing work on bootstraps. Preparing for three-parameter model and testing bootstrap runs. Fixing bugs.

Now running three-parameter bootstraps.

# 5/02/2018

There was an almighty cluster crash on Friday and all the jobs were stopped. Recovered in the afternoon and jobs continuued, but lots of TT108  results missing. Produced a document with limited TT108, but need to run it again.

# 7/02/2018

Meeting with Tomo and John.

John corrected scramble. Scramble is a control, there is also untreated. Do bootstraps on new scramble and untreated.

RAD21 - fit with tau2 = 15 just to see difference. See the separation between pink and red curves (should increase).

Create a table with medians of all bootstraps.

New data for NK1775_ICRF with original (random) order of cells. It is the same data.

Show correlation between tau2 and k2 in RAD21 (from bootstrap).

# 8/02/2018

Made median tables and tau2/k2 plot. Uploading new data.

Fitting scramble for four values of t0. Fitting untreated. And scramble and untreated for 3 pars. Now running bootstraps for 4-par scramble and untreated.

# 9/02/2018

3-par bootstraps started.

# 12/02/2018

3-par bootstraps finished. Updated the document.

# 15/02/2018

Meeting with John.

Scramble had sligthly different definitions of pink and red. There will be new scramble to re-do fits.

Need to update Shiny.


# 15/02/2018

Bootstraps finished. Struggled with the cached version of data, showing the old scramble. Got confused by the difference between shiny app and local document. All sorted out now.

Updating document and Shiny app.
