---
title: "Chromosome compaction - simple three-state model"
output: 
  html_notebook:
    toc: false
    toc_float: false
    code_folding: hide
    css: nice_notebook.css
---

```{r}
library(mylib)
library(ChromCom)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(parallel)
library(latex2exp)
binDir <- "../RData"
```

This is a continuation of the previous notebook, [mod3_notes_old](), with a new model. In the old model, the start time $t_1$ was a fixed number, creating a cusp in the curve, while in the new model is a random variable with exponential distribution.

# 17/08/2017

The new model. An example for $tau = 20$ min, $\Delta t = 0$, $k_1 = 0.04\ {\rm min}^{-1}$ and $k_2 = 0.04\ {\rm min}^{-1}$.

```{r, fig.width=5, fig.height=3}
pars <- c3pars(
  tau = 20,
  dt2 = 0,
  k1 = 0.04,
  k2 = 0.04
)
chr <- ChromCom3(pars)
chr <- generateCells(chr, nsim=1000)
plotTimelines(chr)
```

## Experimental data

```{r, warning=FALSE, fig.width=5}
echr.scr <- experimentalData(dataFile$scramble)
echr.d2 <- experimentalData(dataFile$NCAPD2)
echr.d3 <- experimentalData(dataFile$NCAPD3)
echr.smc <- experimentalData(dataFile$SMC2)
echr.rad <- experimentalData(dataFile$RAD21)
echr.mk <- experimentalData(dataFile$MK1775)
#echr.mki <- experimentalData(dataFile$MK1775_ICRF193)

plotTimelines(echr.scr, smooth=TRUE, k=15)
```



I start with fitting scramble

## Scramble

Fitting with $k_1$, $k_2$ and $tau$ free.

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2")
#chr.scr <- cacheData("fitv_scramble_n10000_tau_k1_k2", fitChr, echr.scr, pars, free, nsim=3000, ntry=32, ncores=6, binDir=binDir)
#plotTimelines(chr.scr, expdata = echr.scr, withpars=TRUE)
```

It looks like we need to free $\Delta t_2$.

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2", "dt2")
chr.scr <- cacheData("fitv_scramble_n10000_tau_k1_k2_dt2", fitChr, echr.scr, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)
plotTimelines(chr.scr, expdata = echr.scr, withpars=TRUE)
```

## NCAPD2

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2", "dt2")
chr.d2 <- cacheData("fitv_NCAPD2_n10000_tau_k1_k2_dt2", fitChr, echr.d2, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)
plotTimelines(chr.d2, expdata = echr.d2, withpars=TRUE)
```


## NCAPD3

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2", "dt2")
chr.d3 <- cacheData("fitv_NCAPD3_n10000_tau_k1_k2_dt2", fitChr, echr.d3, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)
plotTimelines(chr.d3, expdata = echr.d3, withpars=TRUE)
```
## SMC2

```{r, fig.width=5}
pars <- c3pars()
free <- c("tau", "k1", "k2", "dt2")
chr.smc <- cacheData("fitv_SMC2_n10000_tau_k1_k2_dt2", fitChr, echr.smc, pars, free, nsim=10000, ntry=100, ncores=6, binDir=binDir)
#plotTimelines(chr.smc, expdata = echr.smc, withpars=TRUE)
```

# 25/08/2017

Downloaded new data sets that John emailed me a few days ago. New set RAD21 and corrected sets SMC2, MK1775only and MK1775andICRF193.

```{r, fig.width=5}
plotTimelines(chr.smc, expdata = echr.rad, withpars=TRUE)
```

Talking to John and Tomo:


- change $\Delta t_n$ into $\tau_n$, that is, randomize it.
- for scramble add one more param: shift zero. Our definition of zero is not very accurate, so there might be a shift of a few minutes. Start with freezing $t_0$ for scramble.

